---
title: "Introduction to Ultimate Deconvolution"
author: Peter Carbonetto
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
vignette: >
  %\VignetteIndexEntry{Introduction to Ultimate Deconvolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Here we illustrate the features of the Ultimate Deconvolution (UD)
package.

```{r load-pkgs}
library(udr)
```

We simulate 4,000 data points from a UD model with 4 mixture
components:
$$
x_i \sim w_1 N(0,V + U_1) + \cdots + w_4 N(0,V + U_4).
$$

```{r sim-data-settings}
set.seed(1)
n <- 4000
V <- rbind(c(0.8,0.2),
           c(0.2,1.5))
U <- list(none   = rbind(c(0,0),
                         c(0,0)),
          shared = rbind(c(1.0,0.9),
                         c(0.9,1.0)),
          only1  = rbind(c(1,0),
                         c(0,0)),
          only2  = rbind(c(0,0),
                         c(0,1)))
w <- c(0.8,0.1,0.075,0.025)
X <- simulate_ud_data(n,w,U,V)
```

This is the most basic usage, using the default settings:

Note: in this toy simulation example, we set model parameter $V$ to $V$ as we know the simulation truth. In practice, it's likely one needs to estimate $V$ and use $\hat V$ for $V$.

```{r ud-fit-1}
fit0 <- ud_init(X, V = V)
fit1 <- ud_fit(fit0)
```

The default model is a mixture with $K = 10$ mixture components: 2 scaled
scaled and 8 unconstrained prior covariance matrices $U_k$:

```{r ud-fit-2}
summary(fit1)
```

The function `ud_init` is a flexible interface for defining a variety
of UD models. For example,

```{r ud-fit-3}
fit0 <- ud_init(X,U_scaled = U,n_rank1 = 1,n_unconstrained = 1,V = V)
fit2 <- ud_fit(fit0)
summary(fit2)
```

The `ud_fit` interface also allows for flexibly re-fitting a model;
that is, a `ud_fit` output can be used in another call to `ud_fit`.
There is also a more advanced model fitting interface for more
fine-grained control; see `help(ud_fit_advanced)` for details.

Notice that the data do not need to be passed to `ud_fit` because they
are already embedded in the `ud_fit` output. However, one can also
pass a new data set `X` to `ud_fit`.

The `control` argument to `ud_fit` controls the optimization of the
model parameters. For example, the unconstrained prior covariance
matrices are by default optimized using the "truncated eigenvalue
decomposition" (TED) updates, and we can replace these with the
Extreme Deconvolution (ED) updates:

```{r ud-fit-4}
fit3 <- ud_fit(fit0,control = list(unconstrained.update = "ed"))
```

Notice that the likelihood is slightly higher with the TED updates,
reflecting the fact that the TED updates have greater freedom to
optimize the covariance matrices.

These are the current model fitting defaults:

```{r ud-fit-control-default}
unlist(ud_fit_control_default())
```

The most intensive computations are implemented in R (`control$version
= R`) and C++ (`control$version = Rcpp`); both versions should return
the same, or nearly the same output, but the C++ version can sometimes
be much faster.

In some cases each sample might have different measurement error, and
udr can handle such cases by specifying `V` as a list instead of a
matrix:

```{r ud-fit-5}
fit0 <- ud_init(X,V = rep(list(V),n))
fit4 <- ud_fit(fit0)
```

The TED updates cannot be applied to this setting, so the ED updates
must be used instead.

Session info
------------

This is the version of R and the packages that were used to generate
these results:

```{r session-info}
sessionInfo()
```
