---
title: "test general V"
author: "Yunqi Yang"
date: "3/15/2021"
output: html_document
---

```{r  }

set.seed(1)
n <- 500
V <- rbind(c(0.8,0.2),
           c(0.2,1.5))
U <- list(none   = rbind(c(0,0),
                         c(0,0)),
          shared = rbind(c(1.0,0.9),
                         c(0.9,1.0)),
          only1  = rbind(c(1,0),
                         c(0,0)),
          only2  = rbind(c(0,0),
                         c(0,1)))
w <- c(0.8,0.1,0.075,0.025)
X <- simulate_ud_data(n,w,U,V)


```


## compare unconstrained updates for common V vs. general V

Here, I specify a common V and extend it to a list of Vs. fit1 and fit3 both use ``unconstrained.update = "ed"``. In this case, they are doing the same update with the same Vs. So results are expected to be exactly the same. 

However, they are slightly different. The differences occur in the initial log-likelihood. Not sure what caused the error. 

 
```{r  }

set.seed(1)
fit0 <- ud_init(X,V = rep(list(V),n))
fit1 <- ud_fit(fit0,control = list(unconstrained.update = "ed", scaled.update = 'none', rank1.update = 'ed', minval = 0, maxiter = 10 ))

```

```{r  }
set.seed(1)
fit2 <- ud_init(X)
fit3 <- ud_fit(fit2, control = list(unconstrained.update = "ed", scaled.update = 'none', rank1.update = 'none', resid.update = 'none', minval = 0, maxiter = 50 ))


```

```{r  }

plot(fit3$progress$loglik)
points(fit1$progress$loglik)


```

```{r  }

identical(fit0$loglik, fit2$loglik)

```



## compare rank1 updates for common V and general V.

If ``rank1.update = 'ed'``, we are using David's ultimate deconvolution algorithm.
If ``rank1.update = 'teem'``, we are using TEEM rank1 update. 

Those results are not expected to be the same. 
```{r  }

set.seed(1)
fit0 <- ud_init(X,V = rep(list(V),n))
fit1 <- ud_fit(fit0,control = list(unconstrained.update = "none", scaled.update = 'none', rank1.update = 'teem', minval = 0, maxiter = 5000 ))

```


```{r  }

set.seed(1)
fit2 <- ud_init(X)
fit3 <- ud_fit(fit2, control = list(unconstrained.update = "none", scaled.update = 'none', rank1.update = 'teem', resid.update = 'none', minval = 0, maxiter = 5000 ))

```

```{r  }

plot(fit1$progress$loglik)
points(fit3$progress$loglik)

```




